---
title: "Increasing political issue alignment in Europe"
date: 2025-05-19
author:
  - name: Jan Lorenz
    id: jl
    orcid: 0000-0002-5547-7848
    email: jlorenz@constructor.university
    affiliation: 
      - name: Constructor University
        city: Bremen
        state: Germany
        url: www.janlo.de
abstract: |
  This research note shows that issue alignment (as a form of polarization) has almost unanimously increased in European countries from 2002 to 2023 with respect to five variables: left-right selfplacement and attitudes in favor or against of the EU, gays/lesbians, immigrants, and redistribution. This is quantified using principle component analysis of survey responses from the European Social Survey on these five variables for several years and countries. The PC structure is fairly constant over time and also does not vary a lot between different countries. Therefore, the percentage of the variance explained by the first principal components is taken as a proxy for issue aligment. Issue alignment has increased from 29.2% to 35.1%, however, the current level shows that the five issues are by far not fully aligned.  
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
bibliography: references.json
---

```{r}
#| message: false

library(tidyverse)
library(gt)
library(arrow)
library(pals)
library(patchwork)
library(countrycode)

normalize_pc <- function(value) -sign(value[3])*length(value)*value/sum(abs(value))
formatpercentage <- function(x) paste0(round(100*x, digits = 1), "%")
ess <- read_parquet("ESS1-11.parquet")
load("PCAs")
attitudenames <- c("freehms", "gincdif", "lrscale", "impcntr", "euftf")
attitudedesc <- c(#"Homosexual", "Equality", "Political Right", "Immigrants", "EU")
 "Pro gay/lesbian",
 "Pro redistribution",
 "Political right",
 "Pro immigrants",
 "Pro EU")
# These are the countries which have data in all years covered
cntrynames <- c("BE","CH","DE","ES","FI","FR","GB","HU","IE","NL","NO","PL","PT","SE","SI")
```

## Issue alignment

*Issue alignment* is one facet of polarization [@BaldassarriGelman2008PartisansConstraintPolitical] which is about how attitudes on different topic dimensions are alinged. It has also been called attitude consolidation. It means that over time, attitudes become increasingly predictive of each other. When attitudes become correlated, divisions between groups with different attitudes may feel deeper, even in the absence of any change in single-issue polarization on the macro-level. Contrary to single issue based polarization, attitude consolidation does find considerable empirical support [@DellaPosta2020PluralisticCollapseOil] analyzing survey data from the United States of America using belief network analysis. Belief network analysis it is based on the correlation matrix of attitudes computed on survey data. Here, we use a related but arguably simple approach: principle component analysis (PCA).

PCA is a based on the eigendecomposition of the correlation matrix. The eigenvectors are the principle components and the corresponding eigenvalues[^1] represent the part of the total variance explained by the direction the corresponding eigenvector points to. That way the eigenvalues create a natural hierarchy of the principal components. That way, the percentage of total variance explained by the first principal component can serve as a measure of issue alignment. The more variance is explained by the first principle component, the more issues are aligned.

[^1]: All eigenvalues are non-negative because the correlation matrix is symmetric.

## Data

We use five items from the [European Social Survey](https://www.europeansocialsurvey.org/) which are conceptually unrelated: (1) Political selfplacement from left to right, (2) being in favor or against more European unification, (3) being in favor or against that gays and lesbians should live as they want, (4) being for many or few immigrants from poorer countries, and (5) if government should reduce income differences. Conceptually, all these attitudes are independent from each other.[^2] @tbl-items shows the overview of the five variables. Three items scales were reversed such that large numbers match the wording.

[^2]: A left-wing position is traditionally conceptually associated with a preference for reduction of income differences and a right-wing position with a preference against it. Interestingly, it turns out that this association is the lowest on the strongest principal component. Therefore, we can consider it also conceptually independent in modern times.

+:-------------------+:-------------+:-------------------------------------------------------------------------------------+:--------------------------+
| **Belief item**    | **ESS name** | **Item description**                                                                 | **ESS original scale**    |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+
| Political right    | `lrscale`    | Placement on left-right scale                                                        | 0 = Left\                 |
|                    |              |                                                                                      | 10 = Right                |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+
| Pro EU             | `euftf`      | European unification should go further or already gone too far                       | 0 = already gone too far\ |
|                    |              |                                                                                      | 10 = should go further    |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+
| Pro gay/lesbian    | `freehms`    | Gays and lesbians should be free to live life as they wish                           | 1 = Agree strongly\       |
|                    |              |                                                                                      | 5 = Disagree strongly\    |
|                    |              |                                                                                      | (scale reversed)          |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+
| Pro immigrants     | `impcntr`    | Allow many/few immigrants from poorer countries outside Europe to come and live here | 1 = Allow many\           |
|                    |              |                                                                                      | 4 = Allow none\           |
|                    |              |                                                                                      | (scale reversed)          |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+
| Pro redistribution | `gincdif`    | Government should reduce differences in income levels                                | 1 = Agree strongly\       |
|                    |              |                                                                                      | 5 = Disagree strongly\    |
|                    |              |                                                                                      | (scale reversed)          |
+--------------------+--------------+--------------------------------------------------------------------------------------+---------------------------+

: Item overview from the European Social Survey (ESS) {#tbl-items tbl-colwidths="\[15,15,40,30\]"}

For a PCA we need data on all five items for each individual. After filtering for that, data is available for these `r ess |> pull(year) |> unique() |> length()` years: `r ess |> pull(year) |> unique() |> paste(collapse = ", ")`. Further on, we want to make a better comparison over years. To that end we only include countries for which data is available for all these years. These were these `r length(cntrynames)` countries: `r paste0(paste0(countrycode(cntrynames, "iso2c", "country.name"), " (", cntrynames, ")"), collapse = ", ")`. In total we have `r ess |> filter(cntry %in% cntrynames) |> select(all_of(attitudenames)) |> nrow() |> format(big.mark = ",")` individuals in the data set. @tbl-counts shows the number of individual per country and year.

## Main Result: Increasing issue alignment

@fig-main shows how issue alignment increased in almost all European countries for which enough data is available in the European Social survey. The particular structure and stability of the PCA analysis is analyzed in the next sections.  

```{r}
#| label: fig-main
#| fig-cap: Increasing issue alignment measured as explained variance of the first principle componet of country-year specific PCAs based on five variables (Left-right selfplacement and attitude on EU, gay/lesbian, immigrants and redistribution) from the European Social Survey.
#| fig-height: 5.5

ess_y_expVarP1 <- ess_y |> 
	mutate(`PC1 explained Variance` = pca_varexp |> map_dbl(\(x) x[1]),
								cntry = "All") |> 
	select(Year = year, cntry, `PC1 explained Variance` )
ess_y_PC = ess_y |> select(year, pca_pcs) |> unnest(cols = c(pca_pcs))
ess_yc |> select(cntry, year, pca_pcs) |> unnest(cols = c(pca_pcs)) |> 
	filter(word(name, 1) == "PC1") |> 
	mutate(PC = word(name, 1)) |> 
	select(cntry, year, PC, expVar) |> distinct() |> 
	pivot_wider(names_from = year, values_from = expVar) |> 
	na.omit() |>
	pivot_longer(`2004`:`2023`, names_to = "Year", values_to = "PC1 explained Variance") |> 
	mutate(Year = as.numeric(Year)) |> 
	ggplot(aes(Year, `PC1 explained Variance`, color = cntry)) +
	#geom_point() +
	geom_line() +
	geom_point(size = 0.5) +
	geom_line(data = ess_y_expVarP1, linewidth = 1.5, color = "black") +
	geom_point(data = ess_y_expVarP1, size = 2.5, color = "black") +
	#	facet_wrap(~PC, nrow=1) +
	scale_color_manual(values = cols25(16) |> unname()) +
	scale_x_continuous(breaks=c(2004,2006,2008,2012,2014,2016,2018,2020,2023), minor_breaks = c()) +
	scale_y_continuous(labels = scales::percent, limits = c(0.2,0.43)) +
	guides(linewidth = "none", color = guide_legend(title = "Country")) +
	theme_minimal() +
	plot_annotation(caption = "Black thick line = PCAs for all countries")
```

## Principal Compenent Analyses

We compute one PCA for all countries and years, `r nrow(ess_y)` PCAs for each year over all countries, and `r nrow(ess_yc)` PCAs for each country for each year. For visualization purposes we rescale the loadings of the original variables on the different PCs such that the left-right selfplacement is negative. That way, all other variables can be interpreted as being associated with being politically left. However, this only makes sense when the loading for selfplacement is not close to zero. Further on, we make the sum of the absolute values of loading equal to five. 

```{r}
#| label: fig-ess_all
#| fig.cap: "European Social Survey: PCA all years and countries"
#| fig-height: 2

PC_plot <- function(data, title) data |> 
	ggplot(aes(y = topic_desc |> fct_rev(), x = value, fill = value)) + 
	geom_vline(xintercept = c(-1, 1), color = "darkgray") +
	geom_col() +
	scale_fill_gradient2(low = "red", mid = "black", high="green", midpoint = 0, limits = c(-2,2), oob = scales::squish) +
	facet_wrap(~name, ncol = 5) + 
	labs(x = "",y="") +#, title = title) +
	guides(fill = "none") +
	theme_bw()
ess_all |> PC_plot()
```

@fig-ess_all shows the five principal components for the the PCA of the whole data set and the corresponding percentage of the variance explained. PC1 explains 32% of the variance and characterizes a direction where being politically left is associated being in favor of all four other variables. It somehow captures a modern conceptualization of being left and right. PC2 explaining 22.1% of the total variance can be interpreted as a more traditional interpretation, where being left is strongly associated with being in favor of redistribution while there are slightly negative associations with the EU, and immigrants. PC3 distinguishes a direction where being pro gay/lesbian is associated with being against EU und against redistribution. We omit the interpretation of the last two PC's which represent less than 30% of the total variance. 

@fig-ess_y shows the results when PCA's are computed independently for each year. Over all the PC structure remains remarkably stable. Notable changes in loadings are visible for PC2 and PC3. Increasing issue alignment is visble with the explained variance for PC1 increased from 29.2% to 35.1%. 

Finally, @fig-ess-yc shows that PC1 looks very similar for all countries in the dataset in 2023. A notable difference is that former socialist countries (Slovenia, Poland, and Hungary) have no association of left-right selfplacement and being pro or against redistribution. In Portugal the same is true for being pro or against the EU. In comparison, Switzerland (41.8%) and the United Kingdom (39.1%) shows the strongest issue alignment and Portugal (27.6%), Ireland (29.3%) and Slovenia (29.9%) the lowest. 


```{r}
#| label: fig-ess_y
#| fig.cap: "European Social Survey: PCA all countries fro 2004 to 2023."
#| fig-height: 15

g <- map(ess$year |> unique(),
    \(y) ess_y |> filter(year == y) |> pull(pca_pcs) |> first() |> PC_plot() + labs(tag = y))
g[[1]] / g[[2]] / g[[3]] / g[[4]] / g[[5]] / g[[6]] / g[[7]] / g[[8]] / g[[9]]
```

```{r}
#| label: fig-ess_yc
#| fig.cap: "European Social Survey PC1 in countries in 2023."
#| fig-height: 5

ess_yc |> filter(year == 2023, cntry %in% c("BE","CH","DE","ES","FI","FR","GB","HU","IE","NL","NO","PL","PT","SE","SI")) |> 
	select(cntry, year, pca_pcs) |> unnest(cols = c(pca_pcs)) |> 
	bind_rows(ess_y_PC |> filter(year == 2023, ) |> mutate(cntry="All")) |>  
	filter(word(name,1) == "PC1") |>
	mutate(cntry = paste0(cntry, " (", formatpercentage(expVar), ")")) |> 
	ggplot(aes(y = topic_desc |> fct_rev(), x = value, fill = value)) +
	geom_vline(xintercept = c(-1, 1)) +
	geom_col() +
	scale_fill_gradient2(low = "red", mid = "black", high="green", midpoint = 0, limits = c(-2,2), oob = scales::squish) +
	facet_wrap(~cntry) + 
	guides(fill = "none") +
	theme_bw() +
	theme(axis.title.x=element_blank(),axis.title.y=element_blank())
```



## Appendix {.appendix}

```{r}
#| label: tbl-counts
#| tbl-cap: "Number of individuals in dataset per year and country."
 
ess |> filter(cntry %in% cntrynames) |> count(year, cntry) |> 
 pivot_wider(names_from = year, values_from = n) |> 
 rename(Country = cntry) |> 
 knitr::kable()
```
